Plant Modeling Using Environmental Parameters

### 2014-03-28 doc/paper.tex
Start proofreading

L-systems, or Lindenmayer systems, are often used for realistic looking
artificial plant environments. While conceptually simple, L-systems have been
extended with many features to create realistic plants models based on a
description of the plant growth patterns. While significant progress has been
made in providing robust frameworks for modeling artificial plants, the process
of creating specific plants species is still very much a manual process that
often requires a deep understanding of the tools used.

In this project we will explore an alternate approach to plant species creation
by using genetic algorithms to automatically evolve plants suitable to
particular environments.

By tuning environmental parameters for availability of water, the amount of
sunlight, and the ambient soil conditions, artificial plant species are produced
which resembling physical plants that grown in similar environments.

In this approach to plant generation, the underlying L-system production rules
are evolved based on random mutation combined with a fitness function to
determine plant survival rates. In addition to the L-systems, several additional
parameters are introduced to control plant growth and simulation at a more
abstract level.

@@

We closely examined different existing frameworks that we considered using for
our project. The two frameworks that were the best candidates was
OpenAlea\cite{openalea} and Algorithmic Botany\cite{abotany}. Eventually we
decided to implement the software from scratch.

The software was implemented using Java and OpenGL for drawing. The top-level
architecture can be divided into two main categories: plant simulation using
L-systems and the graphical user interface.

@@

The L-systems are implemented in the LSystem and LRule classes, which are
responsible for storing the production rules associated with each plant and
iterating the plants phenotype as the plant grows throughout the simulation. The
Gene and Plant classes store representations of each living plant. The Gene
class represents the plant genotype as a combination of the L-system production
rules along with additional information such as stem and leaf size, stem and
flower colors, the number of seeds produced, and the age at which the plant
matures. Each individual plants phenotype is represented using the Plant class
and contains it's X/Y position and orientation inside the world, along with the
current age and a L-system string (L-string) representing the plants current
shape. The Plant class is also responsible for calculating each individual
plants fitness values at the end of each simulation tick.

@@

Each plant in our environment contains a string of symbol (which we call an
"L-string") representing the plants current shape and size. Each symbol can
represent a stem segment, a flower, a leaf, etc. The plants genotype contains a
set of production rules that govern the growth and development of the plant.
When the plant grows, the production rules are used to replace symbols in the
plants current L-string with sequences of tokens to produces a larger and more
complex plant.

@@

An example of a plant at first generation:

@@

The simulation of the plant environment is governed by ticks. During each tick,
each of the currently living plant objects becomes older. We wanted to have a
similar approach to have plants grow in the real world. Consequently, during the
first 4 ticks the plant grows as a seed under ground. On the fifth tick it
starts growing above ground. When the seventh tick is reached, the plant becomes
mature and grows flowers and leaves, if the plants genotype is capable of it.
The plant object then spreads it seeds within a certain radius. When the season
is over, the plant dies, but it's children grows up with either the same gene,
or a mutation of the gene.

@@

We want to simulate subtle changes in the gene as the mutation isn't suppose to
be extreme in any ways. Gene mutation occurs by choosing one of the production
rules and then mutation the replacement side of the rule. A symbol is randomly
chosen from the replacement to be mutated.

The mutation can either be that the symbols is simply deleted, or that is then
replaced with a sequences from a predefined set of possible mutations.

The replacement sequences we used are logically divided into several logical
groups, however these divisions are just for convenience and are not used by the
mutation algorithm.

@@

For the purpose of fitness calculations, a ``genetic pressure'' version of each
of these parameters is calculated and used in the fitness function.
\[pressure = 1 - (parameter / 100)\]

@@

When calculating each plants fitness several plant parameters are used:

\begin{description}[leftmargin=!,labelindent=0.2in,labelwidth=0.1in]
  \item[Stem ratio]    \hfill \\
    Normalized stem width, $0 .. 1$
  \item[Leaf ratio]    \hfill \\
    Normalized leaf size, $0 .. 1$
  \item[Stem count]    \hfill \\
    The number of stem segments the plant has grown thus far.
  \item[Leaf count]    \hfill \\
    The number of leaves grown thus far.
  \item[Storage space] \hfill \\
    The total storage space available to the plant. We do not calculate an
    actual volume but instead just use a combination of the stem width from the
    plants genes and the number of stem segments.
    \[ storageSpace = stemRatio \times stemCount \]
  \item[Surface area]  \hfill \\
    The total surface area available in the leaves for photosynthesis. Like the
    storage space, this is a combination of the leaf size and the number of
    leaves.
    \[ surfaceArea = leafRatio \times leafCount \]
\end{description}

@@

The plant fitness function is divided by water, sun, an nutrient fitness. Each
of these is calculated separately and the total fitness is the sum. We
implemented the fitness equations such that if all the environmental parameters
are equal, then all the fitness values will also be equal and the plants will
evolve randomly. In a more comprehensive system, some plants could be considered
more or less fit in any environment including an environment with equal
environmental parameters.

@@

\paragraph{Sunlight fitness}
\begin{itemize}
  \item A large surface area collects more light and results in a higher fitness
    in shady environments.
  \item More stem segments will result in a taller plant which can college more
    light. We do not calculate actual heights, but assume that stem segments
    improve light gathering.
\end{itemize}
\vspace{-0.5em}
\begin{eqnarray*}
sunFitness &=& (2 \times surfaceArea \times sunPressure) \\
           &+& (stemCount \times sunPressure)
\end{eqnarray*}

@@

In future programs we would recommend improving the genetic fitness functions by
taking into consideration environmental information related to the neighboring
plants in the environment. This would use more advanced lighting and collision
detection to calculate sunlight based on shadows in the environment.

In addition, more realistic fitness levels could be used that are not linear
with regards to the amount of pressure from different environmental parameters.
In our system, a leafy plant may always win over other plants in a shady
simulation environment even though such as large pant may not be support in a
physical environment to do water or soil limitations.

Some difficulty was encountered evolving plants to appear realistic, and many
limitations are inherit which are difficult to simulation in a computer graphics
setting. For example, two physical plants which share a remarkable resemblances
my be suited to different environments due to small evolutionary changes. Waxy
leaves, for example, could result in a leafy plant well suited to a dry
environment.

>>>>

L-systems, or Lindenmayer systems, are often used to create realistic looking
artificial plant environments. While conceptually simple, L-systems have been
extended with many features to create realistic plants models based on a
description of the plant growth patterns. Significant progress has been made in
providing robust frameworks for modeling artificial plants, but the process of
creating specific plants species is still very much a manual process that often
requires a deep understanding of the tools used.

In this project we will explore an alternate approach to plant species creation
using genetic algorithms to automatically evolve plants suitable to particular
environments.

By tuning environmental parameters for the availability of water, the amount of
sunlight, and the ambient soil conditions, artificial plant species are produced
which resembling physical plants that may grown in similar environments.

With this approach to plant species generation, the underlying L-system
production rules are evolved based on random mutation combined with a fitness
function to determine plant survival rates. In addition to the L-systems,
several additional parameters are introduced to control plant growth and
simulation at a more
abstract level.

@@

We closely examined several existing frameworks that we considered using for
this project. The two frameworks that were the best candidates were Algorithmic
Botany\cite{abotany} and OpenAlea\cite{openalea}.

The Algorithmic Botany suit consists of the VLAB software which can be used on
Mac OS and Linux, the L-studio software for Microsoft Windows, and an easy to
use TreeSketch application for iPad. We primarily experimented with using VLAB
under but found that a significant amount of effort was required to install and
interact with the tools and we estimated that there would be a large learning
curve before we would proficient enough to modify the software for our needs.

OpenAlea is implemented as a set of Python libraries utilizing native code for
certain functions. As with Algorithmic Botany, installation and use of the tools
was challenging. OpenAlea appears to be most targeted towards plant researchers
who wish to manually model specific plants.

As L-systems are conceptually very simple, we decided to create a separate
implementation of the algorithms in order to avoid risks associated with reusing
existing software which we have little experience with. Our software was
implemented using Java and OpenGL for drawing. The top-level architecture can be
divided into two main categories: plant simulation using L-systems and the
graphical user interface.

@@

Our L-systems algorithms are implemented in the LSystem and LRule classes, which
are responsible for storing the production rules associated with each plant and
iterating the plants phenotype as the plant grows throughout the simulation. The
Gene and Plant classes store representations of each living plant. The Gene
class represents the plant genotype as a combination of the L-system production
rules along with additional information such as stem and leaf size, stem and
flower colors, the number of seeds produced, and the age at which the plant
matures. Each individual plants phenotype is represented using the Plant class
and contains it's X/Y position and orientation inside the world, along with the
current age and a L-system string (L-string) representing the plants current
shape. The Plant class is also responsible for calculating each individual
plants fitness values during each simulation tick.

@@

Each plant in our environment contains a string of symbol (which we call an
``L-string'') representing the plants current shape and size. Each symbol can
represent a stem segment, a flower, a leaf, etc. The plants genotype contains a
set of production rules that govern the growth and development of the plant.
When the plant grows, the production rules are used to replace symbols in the
plants current L-string with sequences of tokens to produces a larger and more
complex plant.

@@

Example of a plant at the first generation:

@@

The simulation of the plant environment is governed by ticks. During each tick,
each of the currently living plant objects becomes older. We wanted to have a
similar approach to how plants grow in the real world. Consequently, during the
first 4 ticks the plant grows as a seed under ground. On the fifth tick it
starts growing above ground. When the seventh tick is reached, the plant becomes
mature and grows flowers and leaves, if the plants genotype is capable of it.
The plant object then spreads it seeds within a certain radius. When the season
is over, the plant dies, but it's children grows up with either the same gene,
or a mutation of the gene.

@@

We want to simulate subtle changes in the gene as the mutation isn't suppose to
be extreme in any ways. Gene mutation occurs by choosing one of the production
rules and then mutating the replacement side of the rule. A symbol is randomly
chosen from the replacement to be mutated.

The mutation can either result in a symbol being deleted or begin replaced with
a sequences from a predefined set of possible mutations.

The replacement sequences we used are logically divided into several logical
groups, however these divisions are just for convenience and are not used by the
mutation algorithm:

@@

For the purpose of fitness calculations, a ``genetic pressure'' is calculated
for of each of these parameters and used in the fitness function.
\[pressure = 1 - (parameter / 100)\]

@@

While evaluating the fitness function, several plant parameters are used and
derived parameters are created to simplify the fitness function itself:

\begin{description}[leftmargin=!,labelindent=0.2in,labelwidth=0.1in]
  \item[Stem ratio]    \hfill \\
    Normalized stem width, 0 = narrow stems, 1 = wide stems
  \item[Leaf ratio]    \hfill \\
    Normalized leaf size, 0 = small leaves, 1 = wide leaves
  \item[Stem count]    \hfill \\
    The number of stem segments the plant has grown thus far.
  \item[Leaf count]    \hfill \\
    The number of leaves grown thus far.
  \item[Storage space] \hfill \\
    The total storage space available to the plant. We do not calculate an
    actual volume but instead use a combination of the stem width from the
    plants genes and the number of stem segments.
    \[ storageSpace = stemRatio \times stemCount \]
  \item[Surface area]  \hfill \\
    The total surface area available in the leaves for photosynthesis. Similar
    to storage space, this is a combination of the leaf size and the number of
    leaves.
    \[ surfaceArea = leafRatio \times leafCount \]
\end{description}

@@

The plant fitness function is divided into three parts corresponding to the
water, sun, an nutrient levels. Each of these is calculated separately and the
total fitness is summed afterwards. We implemented the fitness equations such
that if all the environmental parameters are equal, then all the fitness values
will also be equal and the plants will evolve randomly. In a more comprehensive
system, some plants could be considered more or less fit in any environment
including an environment with equal environmental parameters.

@@

\paragraph{Sunlight fitness}
\begin{itemize}
  \item A large surface area collects more light and results in a higher fitness
    in shady environments.
  \item More stem segments will result in a taller plant which can college more
    light. We do not calculate actual heights, but assume that additional stem
    segments improve light gathering.
\end{itemize}
\vspace{-0.5em}
\begin{eqnarray*}
sunFitness &=& (2 \times surfaceArea \times sunPressure) \\
           &+& (stemCount \times sunPressure)
\end{eqnarray*}

@@

In future programs we would recommend improving the genetic fitness functions by
taking into consideration environmental information related to the neighboring
plants. This would require more advanced lighting and collision detection to
calculate sunlight collection and photosynthesis based on shadows in the
environment.

In addition, more realistic fitness levels could be used that are not linear
with regards to the amount of pressure from different environmental parameters.
In our system, a leafy plant may always win over other plants in a shady
simulation environment even though such as large plant may not be support in a
physical environment to do water or soil limitations.

Some difficulty was encountered evolving plants that appear realistic; some
limitations may be inherit to genetic algorithms which rely on graphical
simulations. For example, two physical plants may share a remarkable resemblance
but may be suited to different environments due to small evolutionary changes.
Waxy leaves, for example, could result in a leafy plant well suited to a dry
environment. Additionally, flowering plants often evolve to attract insects and
other animals which act as pollinators and seed dispersers. Without complex
organisms interacting with the plants it may be difficult to evolve certain
features such as fruits and flowers, without predefining certain
characteristics.
